Succinct representations for abstract interpretation

### 2011-10-31 papers/PLDI2012/implicitization.tex
typos

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

@@

Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applied to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

@@

We distinguish three phases in the main loop of the analysis:
\begin{enumerate}
\item \label{step:addingpaths} We start finding a new relevant subset
  $P \cup P'$ of the graph.
  Either the previous iteration or the initialization lead us to a
  state where there are no more paths in the previous subset $P$,
  starting at $p_i$, that make the abstract values of the successors
  grow (otherwise, the SMT solver would not have answered
  ``\emph{unsat}''). Narrowing iterations preserve this property.
  However, there may exist such paths in the entire multigraph, that
  are not in $P$. This phase computes these paths and adds them to
  $P'$. This phase is described in~\ref{subsec:addingpaths}
  and corresponds to lines in~\ref{alg=start-add-paths} to
  \ref{alg=end-add-paths} algorithm~\ref{algo:combined}.
\item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
  $p_i \in P_R$, such that these paths are in $P$, i.e are included in
  the working subgraph. Each time we find a path, we update the
  abstract value of the destination point of the path. This is the
  phase explained in~\ref{subsec:ascending}, and corresponds to
  lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
  algorithm~\ref{algo:combined}.
\item  \label{step:narrowing} We perform narrowing iterations the usual way
  (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
  reiterate from step 1 unless there are no more points to explore,
  i.e. $A' = \emptyset$.
\end{enumerate}

@@

The main loop in the expanded multigraph has then 24 distinct paths, half of
them being self loops, and the other half going to the header of the nested loop.

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_11$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7,8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_11$ with an
		image outside $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_11$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

Our technique gives us the expected invariant 
$\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
Here, only 3 paths out of the 25 have been computed during the analysis. In
practice, depending on the order the SMT-solver returns the paths, other
feasible paths could have been added during the analysis.

@@

	\node[PRstate] (n00) {$p_4$};
	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
	\node[PRstate] (n1) [below of=n0] {$p_{11}$};
	\node (label) [below of=n1, yshift=1.2cm] {Step 1};

@@

	\node[PRstate] (n00) {$p_4$};
	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
	\node[PRstate] (n1) [below of=n0] {$p_{11}$};
	\node (label) [below of=n1, yshift=1.2cm] {Step 2};

@@

In our algorithm, the initialization of the abstract values slightly differs from
algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
\begin{algorithm}[!h]
\begin{algorithmic}[1] 
\FORALL {$k \in \{1,..,m\} \setminus \{\delta\}$}
	\STATE $X_{i,k} \gets \perp$
\ENDFOR
\STATE $X_{i,\delta} \gets I_{p_i}$
\end{algorithmic}
\end{algorithm}

>>>>

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

@@

Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

@@

We distinguish three phases in the main loop of the analysis:
\begin{enumerate}
\item \label{step:addingpaths} We start finding a new relevant subset
  $P \cup P'$ of the graph.
  Either the previous iteration or the initialization lead us to a
  state where there are no more paths in the previous subset $P$,
  starting at $p_i$, that make the abstract values of the successors
  grow (otherwise, the SMT solver would not have answered
  ``\emph{unsat}''). Narrowing iterations preserve this property.
  However, there may exist such paths in the entire multigraph, that
  are not in $P$. This phase computes these paths and adds them to
  $P'$. This phase is described in~\ref{subsec:addingpaths}
  and corresponds to lines in~\ref{alg=start-add-paths} to
  \ref{alg=end-add-paths} in Algorithm~\ref{algo:combined}.
\item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
  $p_i \in P_R$, such that these paths are in $P$, i.e are included in
  the working subgraph. Each time we find a path, we update the
  abstract value of the destination point of the path. This is the
  phase explained in~\ref{subsec:ascending}, and corresponds to
  lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
  Algorithm~\ref{algo:combined}.
\item  \label{step:narrowing} We perform narrowing iterations the usual way
  (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
  reiterate from step 1 unless there are no more points to explore,
  i.e. $A' = \emptyset$.
\end{enumerate}

@@

The main loop in the expanded multigraph has then 8 distinct paths, half of
them being self loops, and the other half going to the header of the nested loop.

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_{10}$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7,8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
		image outside $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_{10}$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

Our technique gives us the expected invariant 
$\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
Here, only 3 paths out of the 8 have been computed during the analysis. In
practice, depending on the order the SMT-solver returns the paths, other
feasible paths could have been added during the analysis.

@@

	\node[PRstate] (n00) {$p_4$};
	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
	\node (label) [below of=n1, yshift=1.2cm] {Step 1};

@@

	\node[PRstate] (n00) {$p_4$};
	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
	\node (label) [below of=n1, yshift=1.2cm] {Step 2};

@@

In our algorithm, the initialization of the abstract values slightly differs from
algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
\begin{algorithm}[!h]
\begin{algorithmic}[1] 
\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
	\STATE $X_{i,k} \gets \perp$
\ENDFOR
\STATE $X_{i,\delta_i} \gets I_{p_i}$
\end{algorithmic}
\end{algorithm}
